Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-08-27 paper.Rnw
made it through the paper with edits! (I may clean up references a bit more.)

The final claim, \textbf{localness}, is addressed by the feature heatmap and metric comparison plot\kge{s}. As stated, the feature heatmap reveal\kge{s}\sout{ed} that the density simulation methods in the examples of this paper result\kge{s}\sout{ed} in global explanations where the same features \kge{are} \sout{were} repeatedly chosen across all (or almost all) observations in the set of explanations. This finding agrees with that of \citep{laugel:2018} who found LIME produced global explanations with the normal approximation simulation method. For the bin based simulation methods in the bullet data example, the feature heatmap showed that the features chosen for the explanations varied between the two classification categories (match versus non-match). This is an interesting finding that suggests that different features can play a role in the predictions of observations in different response categories\kge{.} \sout{, but the patterns of features selected by LIME within a classification category do not vary. Again, this is a suggestion of global explanations.} \kgc{I'm not sure what I was trying to say there.} Furthermore, while the metric comparison plot in the bullet example did not provide agreement between metrics on a best bin based method, all metrics agree that a Gower power of 0.5 for computing the model weights associated with distance of a simulated data point from the prediction of interest is best. This suggests that a less local explanation provided a better explanation of the performance of the random forest.

@@

\kgc{XXX This is where I left off XXX}

@@

The following tuning parameters for the LIME algorithm are available in the \emph{lime} R package.

@@

\item Feature selection method for determining the features to return in an explanation: forward selection applied to a ridge regression, highest weights in ridge regression, LASSO, classification/regression tree splits

@@

Section \ref{exp-scatter} introduces explanation scatterplots under the default simulation method in the \emph{lime} R package: four quantile bins. The structure of an explanation scatterplot remains the same for if any bin based simulation method is used to create the simulated data (any number of quantile or equally spaced bins). However, if the kernel density or normal approximation simulation methods are used as the simulation method, the format of the explanation scatterplot changes. In the density based simulation method scenarios, LIME uses the standardized versions of the predictor variables to fit the explainer model. Thus, the explainer model needs to be represented differently in the explanation scatterplot.

@@

Figure \ref{fig:sine-expl-scat-density} provides example explanation scatterplots for each feature in the default LIME explanation obtained using the kernel density simulation method for the \data \ prediction of interest.

@@

Figure \ref{fig:bullet-expl-scat-density} includes explanation scatterplots for the explanations generated using kernel density simulation for the bullet example cases M and NM discussed in Section \ref{bullet-assess-ex}.

@@

In order to assess the claim of linearity for the \data \ prediction of interest discussed in Section \ref{exp-scatter}, we use one of the most basic diagnostics in a statistician's tool box and draw a residual plot for the explainer model. This is shown in Figure \ref{fig:sine-plot-resids} with the explainer model residuals on the y-axis and explainer model predictions on the x-axis. The points along the x-axes have been jittered to ease the effect of the over-plotted points in the visualization. There is a clear increasing trend in the residuals as the explainer model predictions increase. This is a clear violation of the linearity assumption with the ridge regression model.

@@

For $x_e$ and set of tuning parameters $t$, let $\textbf{X}_{e,t}'$ be the LIME simulated dataset with $K$ features and $S$ observations such that $x'_{e,t,s}$ is the feature vector for simulated data observation $s$. Let $\textbf{Z}'_{e,t}$ be the matrix of interpretability transformed simulated data with $K$ features and $S$ observations such that $z'_{e,t,s}$ is the interpretability transformed feature vector for observation/explanation $e$, tuning parameter $t$, and simulated data observation $s$. Note that $z_{e,t}$ will represent the interpretability transformed version of $x_e$.

Next, let $\omega_t$ represent a proximity distance metric. Then $\omega_t\left(x_e, x'_{e,t,s}\right)$ is the weight assigned to $x'_{e,t,s}$, which is the proximity between $x_e$ and $x'_{e,t,s}$. Allow $g_{e,t}$ to be the explainer model for an explanation $e$ and tuning parameter $t$. Thus, $g_{e,t}\left(z'_{e,s,t}\right)$ is the explainer model prediction for the interpretability transformed simulated data observation $s$.

@@

where $R_{e,t}^2$ is the $R^2$ value for $g_{e,t}$.\\
\\
\emph{Average fidelity} is denoted by $\mathcal{L}_{\mbox{ave}}$ and computed as

\begin{eqnarray*} \mathcal{L}_{\mbox{ave}} & = & \frac{1}{E}\sum_{e=1}^E\mathcal{L}(f, \ g_{e,t}, \ \pi_{t}) \\ & = & \frac{1}{E}\sum_{e=1}^E\sum_{s=1}^{S}\omega_{t}\left(x_e, x'_{e,t,s}\right)\left(f\left(x'_{e,t,s}\right)-g_{e,t}\left(z_{e,t,s}'\right)\right)^2. \end{eqnarray*}\\
\\
\emph{Mean squared explanation error} is denoted by MSEE and computed as

>>>>

The final claim, \textbf{localness}, is addressed by the feature heatmap and metric comparison plot\kge{s}. As stated, the feature heatmap reveal\kge{s}\sout{ed} that the density simulation methods in the examples of this paper result\kge{s}\sout{ed} in global explanations where the same features \kge{are} \sout{were} repeatedly chosen across all (or almost all) observations in the set of explanations. This finding agrees with that of \citet{laugel:2018} who found LIME produced global explanations with the normal approximation simulation method. For the bin based simulation methods in the bullet data example, the feature heatmap showed that the features chosen for the explanations varied between the two classification categories (match versus non-match). This is an interesting finding that suggests that different features can play a role in the predictions of observations in different response categories\kge{.} \sout{, but the patterns of features selected by LIME within a classification category do not vary. Again, this is a suggestion of global explanations.} \kgc{I'm not sure what I was trying to say there.} Furthermore, while the metric comparison plot in the bullet example did not provide agreement between metrics on a best bin based method, all metrics agree that a Gower power of 0.5 for computing the model weights associated with distance of a simulated data point from the prediction of interest is best. This suggests that a less local explanation provided a better explanation of the performance of the random forest.

@@

@@

The following tuning parameters for the LIME algorithm are available in the \emph{lime} R package \citep[pedersen:2020].

@@

\item Feature selection method for determining the features to return in an explanation: forward selection applied to a ridge regression, features with the largest magnitude coefficients in ridge regression, LASSO, classification/regression tree splits

@@

Section \ref{exp-scatter} introduces explanation scatterplots under the default simulation method in the \emph{lime} R package: four quantile bins. The structure of an explanation scatterplot remains the same \sout{for} if any bin based simulation method is used \sout{to create the simulated data}, i.e., any number of quantile or equally spaced bins. However, if the kernel density or normal approximation simulation methods are used \sout{as the simulation method}, the format of the explanation scatterplot changes. In the density based simulation method scenarios, LIME uses the standardized versions of the predictor variables to fit the explainer model. Thus, the explainer model needs to be represented differently in the explanation scatterplot.

@@

Figure \ref{fig:sine-expl-scat-density} provides example explanation scatterplots for each feature in the default LIME explanation obtained using the kernel density simulation method for the \data \ prediction of interest. Figure \ref{fig:bullet-expl-scat-density} includes explanation scatterplots for the explanations generated using kernel density simulation for the bullet example cases M and NM discussed in Section \ref{bullet-assess-ex}.

@@

@@

In order to assess the claim of linearity for the \data \ prediction of interest discussed in Section \ref{exp-scatter}, we use one of the most basic diagnostics in a statistician's tool box and draw a residual plot for the explainer model. This is shown in Figure \ref{fig:sine-plot-resids} with the explainer model residuals on the y-axis and explainer model predictions on the x-axis. The points along the x-axis have been jittered to ease the effect of the over-plotted points in the visualization. There is a clear increasing trend in the residuals as the explainer model predictions increase. This is a clear violation of the linearity assumption with the ridge regression model.

@@

For $x_e$ and \kge{a} set of tuning parameters $t$, let $\textbf{X}_{e,t}'$ be the LIME simulated dataset with $K$ features and $S$ observations such that $x'_{e,t,s}$ is the feature vector for simulated data observation $s$ \kge{corresponding to observation/explanation $e$ and tuning parameters $t$}. Let $\textbf{Z}'_{e,t}$ be the matrix of \kge{simulated data transformed to bin indicator variables (for bin based simulation methods) or standardization (for density based simulation methods)} \sout{interpretability transformed simulated data} with $K$ features and $S$ observations such that $z'_{e,t,s}$ is the interpretability transformed feature vector for observation/explanation $e$, tuning parameter\kge{s} $t$, and simulated data observation $s$. Note that $z_{e,t}$ will represent the \sout{interpretability} transformed version of $x_e$.

Next, let $\omega_t$ represent a proximity distance metric \kge{corresponding to tuning parameters $t$}. Then $\omega_t\left(x_e, x'_{e,t,s}\right)$ is the weight assigned to $x'_{e,t,s}$, which is the proximity between $x_e$ and $x'_{e,t,s}$. Allow $g_{e,t}$ to be the explainer model for an explanation $e$ and tuning parameter $t$. Thus, $g_{e,t}\left(z'_{e,s,t}\right)$ is the explainer model prediction for the interpretability transformed simulated data observation $s$.

@@

\noindent where $R_{e,t}^2$ is the $R^2$ value for $g_{e,t}$.\\
\\
\emph{Average fidelity} is denoted by $\mathcal{L}_{\mbox{ave}}$ and computed as

\begin{eqnarray*} \mathcal{L}_{\mbox{ave}} & = & \frac{1}{E}\sum_{e=1}^E\mathcal{L}(f, \ g_{e,t}, \ \pi_{t}) \\ & = & \frac{1}{E}\sum_{e=1}^E\sum_{s=1}^{S}\omega_{t}\left(x_e, x'_{e,t,s}\right)\left(f\left(x'_{e,t,s}\right)-g_{e,t}\left(z_{e,t,s}'\right)\right)^2. \end{eqnarray*}

\noindent \kge{where $\mathcal{L}$ was originally defined in \citet{ribeiro:2016}.}\\
\\
\emph{Mean squared explanation error} is denoted by MSEE and computed as
